Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-11-15 manuscript.tex
latest

\textit{Manuscript elements}: Figure~1, figure~2, table~1, online
appendices~A and B (including figure~A1 and figure~A2). Figure~2 is to
print in color.

@@

\textit{Keywords}: Examples, model, template, guidelines.

@@

When multiple propagules land on the same territory, they compete to secure the territory as they develop. This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next iteration is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.

@@

Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies; that may in turn change the population mean birth and death rates. This can be seen by summing Eq. \eqref{eq:delttot} over genotypes to get the  change in population size $N$, 
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
\end{equation}
which is independent of $c_i$ (here $L=\sum_j l_j$ is the overalFaltinel propagule density).

@@

The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999}). Stressful environments severely restrict growth and reproduction, so that $b_i\ll 1$ \cite{grime_1974,grime_1977}. Mutations which appreciably improve $b_i$ will be either non-existent or extremely unlikely, so $b_i$ is constrained to remain low. In Grime's view, under these conditions the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate. This implies $b_i/d_i\approx 1$ in our model, and so the propagule density $L$ is suppressed to such low levels that there are essentially no territorial contests occurring.

The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $T$ for a given amount of space. In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense. Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood. In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $T$, not relative to ideal conditions).

@@

In our view, our model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood). In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that resources are spatially localized (e.g. due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle a competitive ability $c$ should be derivable from resource competition. The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It is possible that explicit resource tracking is in fact not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species. We are not aware of any attempts to delineate the conditions under which explicit resource tracking is not required, let alone to connect those results to the density-dependent selection literature [need to check Tilman's book].

On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$. The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to linkage disequilibrium between $c$ on the one hand, and $b$ and $d$ on the other. Previous work on this question in the case of sexual conflict used problem-specific fitness models which directly imposed an absolute fitness cost on competition \citep{rankin_2011,vandoorn_2006}.

In Fig. \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here. In a sense this comparison is unfair: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits). As such, Fig. \ref{fig:table2} is not exhaustive and should be read more as a summary of our model's purpose. Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e. to capture some crucial ecological factors in our most genetically realistic models of evolution. Thus, viewing evolutionary ecology as a spectrum ranging from evolution-only to ecology-only, our model lies closer to the evolution-only end of the spectrum compared to more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology plus mutant invasion --- which lies closer to the ecology-only end of the spectrum.

\begin{figure}
\centering
\begin{tabularx}{\linewidth}{lXXXXX}
&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
  Eco-evo:  \\
  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{55}** & \ding{55} \\
  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
\end{tabularx}
\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology. *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically. Applications to traits come later, and with mixed success (see ``Introduction'' and ``Discussion''). **As a modeling framework, adaptive dynamics is not scenario-specific, and could be applied with any fitness model including our density-dependent lottery, but in practice most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching''.}
\end{figure}

@@

Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$. Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these deviations are negligible unless $m_i$ is very small (say of order $10^2$ or less).

>>>>

\textit{Manuscript elements}:

@@

\textit{Keywords}: r/K selection, absolute fitness, eco-evo, interference competition.

@@

When multiple propagules land on the same territory, they compete to secure the territory as they develop. This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next iteration is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability. This includes cases where territories are won via a sexual contest.

@@

Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies; that may in turn change the population mean birth and death rates. This can be seen by summing Eq. \eqref{eq:delttot} over genotypes to get the  change in population size $N$, 
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
\end{equation}
which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).

@@

The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Stressful environments severely restrict growth and reproduction, so that $b_i\ll 1$ \cite{grime_1974,grime_1977}. Mutations which appreciably improve $b_i$ will be either non-existent or extremely unlikely, so $b_i$ is constrained to remain low. In Grime's view, under these conditions the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate. This implies $b_i/d_i\approx 1$ in our model, and so the propagule density $L$ is suppressed to such low levels that there are essentially no territorial contests occurring.

The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $T$ for a given amount of space. In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense (for empirical support, see \citealt{davis_1998}). Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood. In other words, the stressed population is a high density population where competition is important (that is, high density relative to $T$, not relative to ideal conditions; \citealt{taylor_1990}).

@@

In Fig. \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here. In a sense this comparison is unfair: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits). As such, Fig. \ref{fig:table2} is not exhaustive and should be read more as a summary of our model's purpose. Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e. to incorporate crucial ecological factors in our most genetically realistic models of evolution. Thus, viewing evolutionary ecology \citep{kokko_2007,pelletier_2009,schoener_2011} as a spectrum ranging from evolution-only to ecology-only, our model lies close to the understudied evolution-only end of the spectrum. By comparison, more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology coupled with mutant invasion \citep{diekmann_2004} ---  lie close to the ecology-only end of the spectrum.

\begin{figure}
\centering
\begin{tabularx}{\linewidth}{lXXXXX}
&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
  Eco-evo:  \\
  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{51}** & \ding{55} \\
  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
\end{tabularx}
\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology. *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically. Applications to traits came later, and with mixed success (see ``Introduction'' and ``Discussion''). **In practice, most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' \citep{geritz_1997} or mutualisms \citep{ferriere_2002}, but in principle it can be applied with any fitness model including our density-dependent lottery.}
\end{figure}

In our view, our model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood). In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that resources are spatially localized (e.g. due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition). The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It seems plausible that explicit resource tracking is not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species. We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources. More work is needed connecting resource competition models to the density-dependent selection literature, since most of the existing work is focused on the narrower Grime-Tilman-debate issue of resolving the role of competition at different levels of resource availability \citep{aerts_1999,davis_1998,tilman_2007}.

On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$. The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum fitness trait within a non-zero-sum fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.

@@

Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$. Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these fluctuations are negligible unless $m_i$ is very small (say of order $10^2$ or less). These fluctuations in $m_i$ could be regarded as a feature rather than a flaw since having $m_i$ be exactly constant per generation (for given $b_i$ and $n_i$) is biologically unrealistic. In the canonical model of genetic drift, the Wright-Fisher model, the number of offspring per genotype fluctuates from generation to generation and is approximately Poisson distributed. Nevertheless, for simplicity and ease of comparison with the classic lottery model, we ignore fluctuations in $m_i$ and only account for Poisson fluctuations in the number of propagules landing in each territory.
